---
title: "Week 10 Lab Report"
author: "Antriksh Dhand"
output: pdf_document
date: "2023-05-04"
documentclass: article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Question 3

We throw a fair die $n$ times and study the distribution of the mean of these $n$ throws. Set $n=2$ in this question.

## Part a)

Use the function $\texttt{sample()}$ to simulate $n$ throws. Repeat this $r = 10000$ times, compute the mean of each repetition, and store the results in $\texttt{sm}$.

```{r}
set.seed(0503)

r <- 10000
n <- 2
sm <- rep(0, r)
for (i in 1:r)
{
    x <- sample(1:6, n, replace=TRUE)
    sm[i] <- mean(x)
}

head(sm)
```

## Part b)

Plot the distribution of $\texttt{sm}$ and comment on the shape of the distribution. Does the CLT apply here?

```{r}
plot(table(sm)/r, 
     xlab="values", 
     ylab="density", 
     main="Distribution of the mean")
```

This is not a bell-shaped distribution. In fact the distribution is triangular. The sample size of `n = 2` is much too small for CLT to apply.

## Part c)

Repeat the above procedure for $n = 4, 8, 16, 32$ and comment on how the shape of the distribution changes, relating to the CLT.

```{r}
r <- 10000
n <- c(2, 4, 8, 16, 32, 64)
sm <- rep(0, r)

par(mfrow=c(3, 2))

for (j in n)
{
    for (i in 1:r)
    {
        x <- sample(1:6, j, replace=TRUE)
        sm[i] <- mean(x)
    }
    plot(table(sm)/r, 
         xlab="values", 
         ylab="density", 
         main=paste("n = ", j))
}
```

As the number of coin tosses increases (i.e. an increased number of observations in each repetition), the distribution becomes more and more symmetrical and bell-shaped, resembling a normal curve by the time $n = 16$. This is in line with what the CLT states, which is that as the sample size gets larger, the distribution of sample means $\texttt{sm}$ approximates a normal distribution, regardless of the population's distribution.
