---
title: "MATH1005 Notes"
author: "Antriksh Dhand"
date: "29/11/2021"
output: 
        pdf_document:
                toc: True
                toc_depth: 3
                number_sections: True
---

\newpage

# General R Things

## Reading in data

```{r}
data <- read.csv("WEEK 2/2020fatalities.csv")
```

## Outputting dimension of data

```{r}
dim(data)
```

## Classification of variables (datatypes)

```{r}
str(data)
```

```{r}
# Changing the datatype of specific variables
data$Crash.ID <- as.factor(data$Crash.ID)

# note that you cannot use the levels argument here
# as Month was originally encoded from 1 - 12.
data$Month <- as.factor(data$Month) 
data$Dayweek <- factor(data$Dayweek, 
                       levels = c("Monday", "Tuesday", "Wednesday", 
                                  "Thursday", "Friday", "Saturday", 
                                  "Sunday")) 
# the factor function encodes a vector as a factor (category)
# with order based on the levels argument

# converting to `logical` data format using the batman package
library(batman)
data$Christmas.Period = to_logical(data$Christmas.Period)
data$Easter.Period = to_logical(data$Easter.Period)

str(data)
```

## Difference between `$` and `[]` access operators

```{r}
year <- data$Year
year1 <- data["Year"]
head(year)
head(year1)
```

Using `$` allows you to access individual entries, whereas `[]` returns a vector/dataframe type object.

\newpage

# Data Summaries

## Qualitative data

Often times when working with qualitative data summaries, we need to first produce a frequency table. This can be done using the `table()` function.

```{r}
# outputs a frequency table
table(year)

table(year)[1] # accesses the first element of the frequency table
```

### Bar plot (frequency of 1 categorical variable)

To produce a simple bar plot which counts the frequency of each category, use the `barplot` function which takes in a frequency table as its input.

```{r}
barplot(table(year),
        main = "Number of road fatalities per year",
        xlab = "Year", 
        ylab = "Number of fatalities", 
        ylim = c(0, 3000))
```

\newpage

```{r}
dayweek <- data$Dayweek
barplot(table(dayweek),
        main = "Number of road fatalities by day of the week",
        xlab = "Day of the week",
        ylab = "Number of fatalities",
        ylim = c(0, 10000))
```

\newpage

### Double bar plot (frequency of 2 categorical variables) [Stacked or side-by-side]

Similar to a single bar plot, but now we are comparing against 2 categorical variables.

```{r}
gender <- data$Gender
data1 <- table(gender, dayweek) # in the order that you want the grouping to follow
```

```{r}
# Stacked bar plot
barplot(data1,
        main = "Fatalities by day of the week and gender",
        xlab = "Day of the week",
        ylab = "Number of fatalities",
        ylim = c(0, 10000),
        col = c("red", "lightblue", "lightgreen", "orange"),
        legend = rownames(data1),
        beside = FALSE)
```

\newpage

```{r}
# Side-by-side bar plot
barplot(data1,
        main = "Fatalities by day of the week and gender",
        xlab = "Day of the week",
        ylab = "Number of fatalities",
        ylim = c(0, 7000),
        col = c("red", "lightblue", "lightgreen", "orange"),
        legend = rownames(data1),
        beside = TRUE)
```

\newpage

```{r}
# FILTERING OUT -9 AND UNSPECIFIED
data1_filtered <- data1[-c(1, 4), ] # we don't want rows 1 and 4, but we want all columns

# Side-by-side bar plot but filtered
barplot(data1_filtered,
        main = "Fatalities by day of the week and gender",
        xlab = "Day of the week",
        ylab = "Number of fatalities",
        ylim = c(0, 7000),
        col = c("lightblue", "lightgreen"),
        legend = rownames(data1_filtered),
        beside = TRUE)
```

\newpage

```{r}
speedlimit <- data$Speed.Limit
table(speedlimit)
```

It would make sense to represent speed limit as a categorical variable. We can also introduce a new category of `<40` and merge speeds of 5 up to 30 as they are relatively low frequency.

```{r}
speedlimit[speedlimit == "5"] = "<40"
speedlimit[speedlimit == "10"] = "<40"
speedlimit[speedlimit == "15"] = "<40"
speedlimit[speedlimit == "20"] = "<40"
speedlimit[speedlimit == "25"] = "<40"
speedlimit[speedlimit == "30"] = "<40"
speedlimit[speedlimit == "-9"] = NA

# converting to a factor
speedlimit = factor(speedlimit, 
                    levels = c("<40", "50", "60", "70", "75", "80", "90", 
                               "100", "110", "130"))
table(speedlimit)
```

\newpage

```{r}
# single bar plot
barplot(table(speedlimit),
        main = "Number of fatalities by speed limit",
        xlab = "Speed limit",
        ylab = "Number of fatalities",
        ylim = c(0, 20000))
```

\newpage

```{r}
# double bar plot
data2 = table(gender, speedlimit)
data2_filtered = data2[-c(1, 4), ]
barplot(data2_filtered,
        main = "Number of road fatalities by speed limit and gender",
        xlab = "Speed limit (km/h)",
        ylab = "Number of fatalities",
        col = c("lightblue", "lightgreen"),
        legend = rownames(data2_filtered),
        beside = TRUE)
```

\newpage

## Adjusting axis labels on a chart

There are three new arguments we provide to the charting function here: `las`, `cex.axis`, and `cex.names`.

### `las`

- `las = 0` : both labels are parallel to their axes
- `las = 1` : y-labels are perpendicular, x-labels are parallel
- `las = 2` : both labels are perpendicular to their axes
- `las = 3` : y-labels are parallel, x-labels are perpendicular

### `cex`

`cex` is a function which controls the scaling of text in a graph. 

- `cex.axis` controls the y-axis scaling
- `cex.names` controls the x-axis scaling
- `cex.main` controls the title scaling

```{r}
# Side-by-side bar plot but prettier
barplot(data1_filtered,
        main = "Fatalities by day of the week and gender",
        xlab = "Day of the week",
        ylab = "Number of fatalities", ylim = c(0, 7000),
        col = c("lightblue", "lightgreen"),
        legend = rownames(data1_filtered),
        beside = TRUE,
        las = 1,
        cex.axis = 0.80, cex.names = 0.80, cex.main = 1.5)
```

```{r}
qnorm(.9)
```


\newpage

### Adding lines and points to a graph

We can use the `abline` function to plot lines onto our charts. The line type (`lty`) can be `0` (blank), `1` (solid), `2` (dashed), `3` (dotted) and so on. Some use cases are:

- `abline(a = 1, b = 2, col = 'blue', lty = 2)` : plots a blue dashed line with $\text{intercept} = 1$ and $\text{slope} = 2$ 
- `abline(h = 4, col = 'green', lty = 1)` : plots a horizontal green solid line at $y = 4$
- `abline(v = mean(1, 2, 3, 4, 5))` : plots a vertical black solid line at $x = 3$

You can also use `lwd` (line width) to control the thickness of the lines.

```{r}
data_houses <- read.csv("WEEK 3/NewtownJune2017.csv")
hist(data_houses$Sold, 
     main = "Newtown properties", 
     xlab = "Price (in 1000s)")
abline(v = mean(data_houses$Sold), col = "red")
abline(v = median(data_houses$Sold), col = "purple", lty = 2, lwd = 3)
```

\newpage

## Quantitative data

### Histogram (frequency of 1 quantitative variable)

```{r}
data$Age[data$Age == -9] <- NA # cleaning

age <- data$Age
hist(age)
```

Note that this histogram is produced on the frequency scale. We prefer to use the density scale.

\newpage

```{r}
# choosing specific class intervals
par(mfrow = c(1, 2))
hist(age,
     br = c(0, 18, 25, 70, 101), # custom class intervals
     freq = FALSE, # produces the histogram on the density scale
     right = F, # makes the intervals right-open
     xlab = "Age (in years)",
     ylab = "Density")

hist(age,
     br = 10,
     freq = FALSE, # produces the histogram on the density scale
     right = F, # makes the intervals right-open
     xlab = "Age (in years)",
     ylab = "Density")
```

\newpage

### Multiple histograms (frequency of 1 quantitative variable split by a categorical variable)

```{r}
ageF <- data$Age[data$Gender == "Female"] # selecting just the female ages
ageM <- data$Age[data$Gender == "Male"]

par(mfrow = c(1, 2)) # outputting in 1 row and 2 columns
hist(ageF, freq = F)
hist(ageM, freq = F)
```

\newpage

### Single boxplot (numerical summary of a quantitivate variable)

```{r}
summary(age)
boxplot(age,
        main = "Boxplot of ages involved in fatalities",
        horizontal = TRUE)
```

\newpage

### Multiple boxplots (numerical summary of a quantitivate variable split up by a qualitative variable)

A comparative boxplot splits up a quantitative variable by a qualitative variable.

```{r}
summary(age[gender == "Female"])
summary(age[gender == "Male"])
```

```{r}
gender <- factor(gender, levels = c("Female", "Male"))
boxplot(age ~ gender, horizontal = TRUE) # plot gender split by age
```

\newpage

### Scatter plots (compare two quantitative variables)

```{r}
# plot()
```

### Heatmaps

A heatmap is especially useful when a contingency table is not practical due to too many different values. We can also scale (centered and scaled in a specific direction), the heatmap by column — results in something like an Age histogram for every year.

```{r}
par(mfrow = c(1, 2))
heatmap(table(age, year), 
        Rowv = NA, 
        Colv = NA, 
        scale = "none", 
        main = "Heatmap of frequencies")

# scaled by column
heatmap(table(age, year), 
        Rowv = NA, 
        Colv = NA, 
        scale = "col", 
        main = "Heatmap scaled by column")

```

\newpage

# Numerical summaries: Centre

## Sample mean

- The average of the data
- Involves all the data
- Serves as a balancing point

$$\bar{x} = \frac{\sum_{i=1}^n x_i}{n}$$

In R we just use the `mean()` function.

```{r}
mean(data_houses$Sold)
```

## Sample median

The middle data point when the observations are ordered from smallest to largest.

For odd $n$:
$$\tilde{x} = x_{(\frac{n+1}{2})}$$

For even $n$:
$$\tilde{x} = \frac{x_{\left( \frac{n}{2} \right)} + x_{\left( \frac{n}{2} + 1 \right)}}{2}$$

In R we use the `median()` function.

```{r}
median(data_houses$Sold)
```


```{r}
hist(data_houses$Sold)
abline(v = mean(data_houses$Sold), col = "green", lwd = 2)
abline(v = median(data_houses$Sold), col = "purple", lwd = 2)
```

The sample median is said to be robust and is a good summary for skewed data as it is not affected by outliers.

# Numerical summaries: Spread

## Sample SD
$$\text{SD}_\text{sample} = \sqrt{\frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n-1}}$$

In R, the `sd` function **always calculates the sample standard deviation.**

You can convert from sample SD to population SD by a conversion factor:
$$\text{SD}_\text{pop} = \text{SD}_\text{sample} \, \times \sqrt{\frac{n-1}{n}}$$

## Population SD
$$\text{SD}_\text{pop} = \sqrt{\frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n}}$$
In R, calculating the population SD requires the `multicon` package.

```{r}
library(multicon)
popsd(data_houses$Sold)
```


## Z-Score
Measures how many SD's a point is above or below the mean.

$$z = \frac{x - \bar{x}}{\text{SD}}$$

## IQR
Range of the middle 50% of the data.
$$\text{IQR} = Q_3 - Q_1$$

## Lower and upper thresholds
$$\text{LT} = Q_1 - 1.5 \cdot \text{IQR}$$
$$\text{UT} = Q_3 + 1.5 \cdot \text{IQR}$$

## Coefficient of variation
$$\text{CV} = \frac{\text{SD}}{\bar{x}}$$


# General Normal Curve

Suppose the heights of Australian women follow a normal distribution with mean 161.8cm and SD 7.7cm. What proportion of women will have height less than 170cm?

We can use the `pnorm(x, mean, SD, lower.tail = TRUE/FALSE)` function to evaluate this.

```{r}
pnorm(170, 161.8, 7.7, lower.tail = TRUE) 
```

# Scatter plots and correlation

A scatter plot can be summarised by 5 numerical summaries:

- sample mean and sample SD of X $(\bar{x}, \text{SD}_x)$
- sample mean and sample SD of Y $(\bar{y}, \text{SD}_y)$
- correlation coefficient ($r$)

## Correlation coefficient

Measures the clustering of the data around a line. It is the mean of the product of the variables in standard units.

$$r = \frac{1}{n} \sum_{i=1}^n \frac{x_i - \bar{x}}{\text{SD}_x} \frac{y_i - \bar{y}}{\text{SD}_y}$$

```{r}
data_class <- read.csv("WEEK 5/LAB/math1005.csv")
```

```{r}
# data cleaning (removing na values)
height <- na.omit(data_class$Height)[1:358]
shoesize <- na.omit(data_class$ShoeSize)[1:358]
```

```{r}
plot(height, shoesize)
```
```{r}
# further data cleaning 
# (removing any entries where the shoe size is greater than 20)
height <- height[shoesize <= 20]
shoesize <- shoesize[shoesize <= 20] 
```

```{r}
plot(height, shoesize,
     main = "Scatter plot of shoe sizes versus height",
     xlab = "Height (cm)",
     ylab = "Shoe size")
```

```{r}
library(multicon)
SU_x = (height - mean(height))/popsd(height) # converting to standard units
SU_y = (shoesize - mean(shoesize))/popsd(shoesize)
mean(SU_x * SU_y) # mean of the product of the variables in standard units
```

In R, we can use the `cor(x, y)` function to calculate the correlation coefficient.
```{r}
cor(SU_x, SU_y)
```


## Regression lines

### SD line (bad descriptor)

The SD line connects the points of averages $(\bar{x}, \bar{y})$ to $(\bar{x} + \text{SD}_x,\, \bar{y} + \text{SD}_y)$. As it does not use the correlation coefficient, it is insensitive to the amount of clustering around the line. It generally under- and over-estimates at the extremes.

### The regression line

To completely describe the scatter plot, we need to use all 5 numerical summaries: $\bar{x}$, $\bar{y}$, $\text{SD}_x$, $\text{SD}_y$ and $r$.

The **regression line** connects $(\bar{x}, \bar{y})$ to $(\bar{x} + \text{SD}_x,\, \bar{y} + r\text{SD}_y)$. The $r$ term in the second point's $y$-coordinate controls the steepness and the sign of the gradient. It improves the fit of the line at the extremes. Hence, the slope of the regression line is
$$m = \frac{(\bar{y} + r\text{SD}_y) - \bar{y}}{(\bar{x} + \text{SD}_x) - \bar{x}} = r \cdot \frac{\text{SD}_y}{\text{SD}_x}$$
The y-intercept of the regression line is
$$b = \bar{y} - m\bar{x}$$
In R, we use the `lm(y ~ x)` function (linear model) to find the regression line. The input is `y ~ x` as in "$y$ variable regressed onto $x$ variable".

```{r}
linear_model <- lm(shoesize ~ height)
print(linear_model)
```

The `lm` function has many callable parameters such as `$coefficients` which returns a named vector of coefficients where `[1]` is the intercept and `[2]` is the slope.

```{r}
linear_model$coefficients
```

So the regression line for this linear model is $y = 1.33 + 0.0445x$. To plot this regression line onto the scatter plot we can use `abline()`.

```{r}
plot(height, shoesize,
     main = "Scatter plot of shoe sizes versus height",
     xlab = "Height (cm)",
     ylab = "Shoe size")
abline(lm(shoesize ~ height), col = "red")
```

The best method for making predictions is the regression line as it includes all the data. You can use the regression line to predict $y$ from $x$ or $x$ from $y$ but you **must refit the data**: you cannot simply rearrange the formula.

## Residuals

A residual is the vertical distance of a point above or below the regression line. It represents the error between the actual value and the prediction.
$$e_i = y_i - \hat{y}_i$$
where $y_i$ is the actual value and $\hat{y}_i$ is the prediction.

### Residual plot

- A residual plot graphs the residuals vs $x$.
- The residual plot is a diagnostic plot to check the appropriateness of a linear model.
- If the linear fit is appropriate for the data, it should show no pattern (random about $y = 0$).

In R, the residuals are built into the linear model.

```{r}
plot(height, linear_model$residuals, ylab = "Residuals", xlab = "Height")
abline(h = 0)
```

- If there is the same (equal) variance around the y axis, the data displays *homoscedasticity* and the data is appropriate to be fit using a linear model. The regression line can be used for predictions. 
- If not, the data displays *heteroscedasticity* and the data should not be fit using a linear model.

# Chance

## Simulating chance

Write a simulation for a game where you a toss a die 4 times and you win if you get at least one 1.

```{r}
gameA <- function() {
        rolls <- sample(1:6, size = 4, replace = TRUE)
        condition <- sum(rolls == 1) > 0
        return(condition)
}

simsA <- replicate(1e+05, gameA()) # results in a table of T/F values if won/lost
probability <- sum(simsA)/length(simsA) # calculates the probability of winning
probability
```



